---
title: "Criterion Continuous Concentration (CCC) Derivation for Microcystins in California Freshwater Aquatic Species"
subtitle: "EPA (1985) Guidelines methodology"
author: "SCCWRP, Leah Thornton Hampton"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    df-print: kable
  docx:
    toc: true
execute:
  echo: true
  warning: false
---

## Overview

This document reproduces, in R, the Criterion Continuous Concentration (CCC)
derivation performed for the microcystin-LR toxicity dataset
(`ToxSummary.csv`), following the general approach in the US EPA (1985)
*Guidelines for Deriving Numerical National Water Quality Criteria for the
Protection of Aquatic Organisms and Their Uses* (Stephan et al. 1985).

Four scoping decisions drive every calculation below, applied in this
order:

1. **Endpoint filter (applied first, all data):** only rows where
   `endpoint` is one of *Growth, Development, Reproduction, Mortality,* or
   *Growth + Development* are retained — i.e., whole-organism/apical
   endpoints. Behavior, Histology, and Subcellular (gene/protein) endpoints
   are excluded.
2. **Species scope:** only rows flagged `ca_relevant == "Yes"` are used.
3. **Censored values:** any test reporting only a NOEC (no LOEC) or only a
   LOEC (no NOEC) is excluded, since the standard EPA Chronic/Acute Value
   requires both bounds.
4. **Acute side further restricted to Mortality (Step 5):** on top of the
   Step 1 endpoint filter, the Acute Value calculation uses only rows with
   `endpoint == "Mortality"`, consistent with EPA guidance that Final
   Acute Values be based on severe/lethal-type endpoints rather than
   sublethal ones. The chronic side is unaffected and still uses all four
   endpoints (Growth, Development, Reproduction, Mortality) from Step 1.

::: {.callout-important}
## Headline conclusion: this dataset does not support an EPA-guideline-compliant CCC

EPA's guidelines require toxicity data spanning a minimum of **8 taxonomic
categories** (salmonid, a second/warmwater fish family, a third chordate,
planktonic crustacean, benthic crustacean, insect, a non-arthropod/
non-chordate phylum, and one more family) before either a Final Chronic
Value (direct method) or a Final Acute Value (used in the FAV/FACR
fallback method) can be treated as validly calculated.

With the filters above applied, the **chronic** subset covers only 2 genera
/ 2 categories, and the **acute** subset covers only 6 genera / 3
categories. **Neither the direct method nor the FAV/FACR fallback clears
EPA's minimum data bar with this data.** The fallback method only rescues
the calculation if the acute data are *more* taxonomically complete than
the chronic data — here, they aren't complete either.

The CCC calculated below (≈0.0277 µg/L) should therefore be read strictly
as **a worked demonstration of the calculation mechanics**, not a
defensible criterion. It is also highly sensitive to scope: an earlier
version of this same calculation with a broader apical-endpoint filter
(Growth/Development/Reproduction/Mortality on both acute and chronic sides)
produced CCC ≈ 0.0235 µg/L, and an even earlier version with no endpoint
filter at all produced CCC ≈ 1.34 µg/L — nearly a 60-fold range across the
three versions, driven entirely by which endpoints were allowed in. That
sensitivity is itself a symptom of too little independent data to average
out any one study's influence.

Additionally, the exact Final Acute Value formula was reconstructed from
established technical literature/prior knowledge rather than freshly
re-verified against a primary source in this session — see the note below
Step 5.
:::

## Setup

No external packages are required — everything here uses base R so the
script runs anywhere R is installed.

```{r setup}
geomean <- function(x) exp(mean(log(x)))

parse_genus <- function(species_col) {
  m <- regmatches(species_col, regexpr("\\(([^)]+)\\)", species_col))
  binom <- trimws(gsub("[()]", "", m))
  genus <- sapply(strsplit(binom, "\\s+"), `[`, 1)
  data.frame(binomial = binom, genus = genus, stringsAsFactors = FALSE)
}

KEEP_ENDPOINTS <- c("Growth", "Development", "Reproduction", "Mortality",
                    "Growth + Development")
```

## Step 1 — Load data and parse genus

```{r load-data}
df <- read.csv("Data/ToxSummary.csv", stringsAsFactors = FALSE)
df$noec_ug_l <- as.numeric(df$noec_ug_l)
df$loec_ug_l <- as.numeric(df$loec_ug_l)

parsed <- parse_genus(df$Species)
df$binomial <- parsed$binomial
df$genus    <- parsed$genus

nrow(df)
table(df$endpoint)
```

## Step 2 — Apply the endpoint filter (upfront, before any other filter)

```{r endpoint-filter}
df_ep <- df[df$endpoint %in% KEEP_ENDPOINTS, ]
cat("Rows before endpoint filter:", nrow(df), "\n")
cat("Rows after endpoint filter: ", nrow(df_ep), "\n")
table(df_ep$endpoint)
```

## Step 3 — Apply the scoping filter (`ca_relevant == "Yes"`)

```{r scope-filter}
ca <- df_ep[trimws(tolower(df_ep$ca_relevant)) == "yes", ]
nrow(ca)
```

## Step 4 — Chronic side: Chronic Value, SMCV, GMCV

```{r chronic-chv}
chronic_all    <- ca[trimws(tolower(ca$chronic_acute)) == "chronic", ]
chronic_usable <- chronic_all[!is.na(chronic_all$noec_ug_l) &
                                 !is.na(chronic_all$loec_ug_l), ]
chronic_excluded <- chronic_all[is.na(chronic_all$noec_ug_l) |
                                   is.na(chronic_all$loec_ug_l), ]

chronic_usable$ChV <- sqrt(chronic_usable$noec_ug_l * chronic_usable$loec_ug_l)

cat("Chronic rows after endpoint+ca_relevant filters:", nrow(chronic_all), "\n")
cat("Usable (both NOEC & LOEC present):", nrow(chronic_usable), "\n")
cat("Excluded (censored, one bound missing):", nrow(chronic_excluded), "\n")

chronic_usable[, c("Species", "reference", "endpoint",
                    "noec_ug_l", "loec_ug_l", "ChV")]
```

**Species Mean Chronic Value (SMCV)** = geometric mean of ChV within a species.
**Genus Mean Chronic Value (GMCV)** = geometric mean of SMCV within a genus.

```{r chronic-smcv-gmcv}
smcv <- aggregate(ChV ~ binomial + genus, data = chronic_usable, FUN = geomean)
names(smcv)[names(smcv) == "ChV"] <- "SMCV"

n_tests <- aggregate(ChV ~ binomial, data = chronic_usable, FUN = length)
names(n_tests)[2] <- "n_tests"
smcv <- merge(smcv, n_tests, by = "binomial")

gmcv <- aggregate(SMCV ~ genus, data = smcv, FUN = geomean)
names(gmcv)[names(gmcv) == "SMCV"] <- "GMCV"
gmcv <- gmcv[order(gmcv$GMCV), ]

smcv
gmcv
```

### Minimum data requirement check (chronic)

EPA requires chronic data for a minimum of 8 families spanning specific
taxonomic categories before a Final Chronic Value can be calculated
directly.

```{r chronic-groups}
unique(chronic_usable[, c("genus", "group")])
```

Only `r nrow(gmcv)` genera / `r length(unique(chronic_usable$group))`
taxonomic categories are represented — well short of the 8-family minimum.
**A direct Final Chronic Value cannot be validly calculated from this
subset**, so the script proceeds with EPA's FAV/FACR fallback method (with
the caveat, confirmed in Step 6 below, that the acute side doesn't clear
the bar either).

### Variant: what if censored (unbounded) NOEC/LOEC rows are included?

The calculation above excludes any chronic row missing either bound, since
a Chronic Value is properly defined as $\sqrt{\text{NOEC}\times\text{LOEC}}$
— a one-sided ("censored") row only tells you the true threshold is
*somewhere above* a NOEC-only value, or *somewhere below* a LOEC-only
value; it isn't itself a Chronic Value. This variant asks: what if we use
that single available bound as a stand-in for the Chronic Value anyway, to
see whether it changes the taxonomic-sufficiency picture?

Doing this is **not** standard EPA practice, and it introduces a
direction-specific bias: substituting a bare NOEC understates the true
Chronic Value (overly conservative/protective), while substituting a bare
LOEC overstates it (under-protective). Mixing both types into one
geometric mean, as done below, is a further deviation on top of that. This
section is included purely as a documented sensitivity check, not as an
input to the main CCC calculation that follows.

```{r chronic-variant}
# Surrogate ChV: use sqrt(NOEC*LOEC) where both are present, otherwise
# fall back to whichever single bound is available.
chronic_all$ChV_surrogate <- ifelse(
  !is.na(chronic_all$noec_ug_l) & !is.na(chronic_all$loec_ug_l),
  sqrt(chronic_all$noec_ug_l * chronic_all$loec_ug_l),
  ifelse(!is.na(chronic_all$noec_ug_l), chronic_all$noec_ug_l,
         ifelse(!is.na(chronic_all$loec_ug_l), chronic_all$loec_ug_l, NA))
)
chronic_all$bound_type <- ifelse(
  !is.na(chronic_all$noec_ug_l) & !is.na(chronic_all$loec_ug_l), "both",
  ifelse(!is.na(chronic_all$noec_ug_l), "NOEC-only (censored above)",
         ifelse(!is.na(chronic_all$loec_ug_l), "LOEC-only (censored below)",
                "neither")))

chronic_variant <- chronic_all[!is.na(chronic_all$ChV_surrogate), ]

cat("Genera with a usable value, both-required version:", length(unique(chronic_usable$genus)), "\n")
cat("Genera with a usable value, censored-included variant:", length(unique(chronic_variant$genus)), "\n")

chronic_variant[, c("Species", "genus", "group", "bound_type", "ChV_surrogate")]
```

```{r chronic-variant-smcv-gmcv}
smcv_variant <- aggregate(ChV_surrogate ~ binomial + genus, data = chronic_variant, FUN = geomean)
names(smcv_variant)[names(smcv_variant) == "ChV_surrogate"] <- "SMCV"

gmcv_variant <- aggregate(SMCV ~ genus, data = smcv_variant, FUN = geomean)
names(gmcv_variant)[names(gmcv_variant) == "SMCV"] <- "GMCV"
gmcv_variant <- gmcv_variant[order(gmcv_variant$GMCV), ]

smcv_variant
gmcv_variant
```

```{r chronic-variant-groups}
unique(chronic_variant[, c("genus", "group")])
```

Including the censored rows brings the chronic side from `r nrow(gmcv)`
genera up to `r nrow(gmcv_variant)` genera — `r nrow(gmcv_variant)` is
exactly the bare minimum the FCV formula needs to run at all (it requires
at least 4 GMCVs). But taxonomic **category** coverage barely moves: the
newly-added genus (*Salmo*) falls in the same Salmonid category as
*Oncorhynchus*, which was already represented, so category coverage stays
at `r length(unique(chronic_variant$group))` of the 8 required — no closer
to EPA's minimum than before, just barely enough genera to make the
arithmetic run.

```{r chronic-variant-fcv}
N_variant <- nrow(gmcv_variant)

if (N_variant >= 4) {
  gmcv_variant$R      <- seq_len(N_variant)
  gmcv_variant$P      <- gmcv_variant$R / (N_variant + 1)
  gmcv_variant$lnGMCV <- log(gmcv_variant$GMCV)
  gmcv_variant$sqrtP  <- sqrt(gmcv_variant$P)

  four_v <- head(gmcv_variant, 4)

  sum_lnGMCV_v  <- sum(four_v$lnGMCV)
  sum_lnGMCV2_v <- sum(four_v$lnGMCV^2)
  sum_P_v       <- sum(four_v$P)
  sum_sqrtP_v   <- sum(four_v$sqrtP)

  S2_v  <- (sum_lnGMCV2_v - (sum_lnGMCV_v^2) / 4) / (sum_P_v - (sum_sqrtP_v^2) / 4)
  S_v   <- sqrt(S2_v)
  L_v   <- (sum_lnGMCV_v - S_v * sum_sqrtP_v) / 4
  A_v   <- S_v * sqrt(0.05) + L_v
  FCV_variant <- exp(A_v)

  data.frame(quantity = c("N genera", "S^2", "S", "L",
                           "A = S*sqrt(0.05)+L", "FCV (variant, ug/L)"),
             value = c(N_variant, S2_v, S_v, L_v, A_v, FCV_variant))
} else {
  FCV_variant <- NA
  cat("Only", N_variant, "genera -- still cannot run the 4-point formula.\n")
}
```

```{r chronic-variant-sanity}
if (!is.na(FCV_variant)) FCV_variant < min(gmcv_variant$GMCV)
```

**Takeaway:** this variant produces FCV ≈ `r if (!is.na(FCV_variant)) round(FCV_variant, 3) else NA` µg/L
and technically clears the bare mathematical minimum (4 GMCVs) — but it
does so only by mixing in one-sided censored values with an unresolved
directional bias, and it makes no progress on the taxonomic-category
shortfall that is the actual reason this dataset can't support a
guideline-compliant Final Chronic Value. It is not used in the CCC
calculation that follows; the main pipeline continues with the
both-bounds-required `chronic_usable`/`smcv`/`gmcv` objects from above.

## Step 5 — Acute side: Acute Value, SMAV, GMAV

```{r acute-av}
acute_all <- ca[trimws(tolower(ca$chronic_acute)) == "acute", ]

# Acute side is further restricted to the Mortality endpoint only (on top
# of the Step 2 endpoint filter). Both conditions must be combined into a
# single row-index expression with `&`/`|` -- putting a condition in the
# *column* position (df[rows, cols]) silently does the wrong thing instead
# of throwing an error, so it's an easy mistake to make.
acute_usable <- acute_all[!is.na(acute_all$noec_ug_l) &
                             !is.na(acute_all$loec_ug_l) &
                             acute_all$endpoint == "Mortality", ]
acute_excluded <- acute_all[is.na(acute_all$noec_ug_l) |
                               is.na(acute_all$loec_ug_l) |
                               acute_all$endpoint != "Mortality", ]

# Surrogate Acute Value = geometric mean of NOEC & LOEC (a true LC50/EC50
# was not consistently reported in this dataset; flagged as a deviation
# from strict EPA practice in the report).
acute_usable$AV <- sqrt(acute_usable$noec_ug_l * acute_usable$loec_ug_l)

cat("Acute rows after endpoint+ca_relevant filters:", nrow(acute_all), "\n")
cat("Usable (Mortality endpoint, both NOEC & LOEC present):", nrow(acute_usable), "\n")
cat("Excluded (non-Mortality endpoint, or censored):", nrow(acute_excluded), "\n")
```

```{r acute-smav-gmav}
smav <- aggregate(AV ~ binomial + genus, data = acute_usable, FUN = geomean)
names(smav)[names(smav) == "AV"] <- "SMAV"

gmav <- aggregate(SMAV ~ genus, data = smav, FUN = geomean)
names(gmav)[names(gmav) == "SMAV"] <- "GMAV"
gmav <- gmav[order(gmav$GMAV), ]

N_acute <- nrow(gmav)
gmav$R      <- seq_len(N_acute)
gmav$P      <- gmav$R / (N_acute + 1)
gmav$lnGMAV <- log(gmav$GMAV)
gmav$sqrtP  <- sqrt(gmav$P)

smav
gmav
```

### Minimum data requirement check (acute)

```{r acute-groups}
unique(acute_usable[, c("genus", "group")])
```

Only `r N_acute` genera / `r length(unique(acute_usable$group))` taxonomic
categories are represented on the acute side too — also short of the
8-family minimum (missing, at least, a salmonid, a second fish family, and
a benthic crustacean). **This means the FAV/FACR fallback does not actually
rescue this derivation** from EPA's data-sufficiency requirement — it is
undermined by the same kind of taxonomic gap as the direct chronic method.

## Step 6 — Final Acute Value (FAV)

EPA's method: rank the Genus Mean Acute Values (GMAVs) ascending, assign
cumulative probability $P = R/(N+1)$, take the four lowest GMAVs, and fit a
log-triangular extrapolation to the 5th percentile ($P = 0.05$):

$$
S^2 = \frac{\sum(\ln \text{GMAV})^2 - \frac{(\sum \ln \text{GMAV})^2}{4}}
           {\sum P - \frac{(\sum \sqrt P)^2}{4}}
\qquad
S = \sqrt{S^2}
\qquad
L = \frac{\sum \ln(\text{GMAV}) - S \sum \sqrt P}{4}
\qquad
\text{FAV} = e^{\,S\sqrt{0.05} + L}
$$

```{r fav-calc}
four <- head(gmav, 4)
four

sum_lnGMAV  <- sum(four$lnGMAV)
sum_lnGMAV2 <- sum(four$lnGMAV^2)
sum_P       <- sum(four$P)
sum_sqrtP   <- sum(four$sqrtP)

S2  <- (sum_lnGMAV2 - (sum_lnGMAV^2) / 4) / (sum_P - (sum_sqrtP^2) / 4)
S   <- sqrt(S2)
L   <- (sum_lnGMAV - S * sum_sqrtP) / 4
A   <- S * sqrt(0.05) + L
FAV <- exp(A)

data.frame(quantity = c("N genera", "sum(lnGMAV)", "sum(lnGMAV^2)",
                         "sum(P)", "sum(sqrtP)", "S^2", "S", "L",
                         "A = S*sqrt(0.05)+L", "FAV"),
           value = c(N_acute, sum_lnGMAV, sum_lnGMAV2, sum_P, sum_sqrtP,
                     S2, S, L, A, FAV))
```

**Sanity check:** a valid 5th-percentile extrapolation should fall at or
below the single lowest GMAV in the dataset.

```{r fav-sanity-check}
FAV < min(gmav$GMAV)
```

**Formula verification note:** this formula was reconstructed from
established technical literature/prior knowledge; live retrieval of the
primary EPA 1985 Guidelines PDF in the session that produced this analysis
returned inconsistent, garbled formula text (a PDF text-extraction
artifact), so the exact algebra above was not freshly re-verified against
the primary source. The sanity check above passes, but please check
against Stephan et al. (1985), Appendix I, before relying on this number.

## Step 7 — Acute-Chronic Ratio (ACR) and Final ACR (FACR)

```{r acr-facr}
acr <- merge(smav[, c("binomial", "genus", "SMAV")],
             smcv[, c("binomial", "SMCV")], by = "binomial")
acr$ACR <- acr$SMAV / acr$SMCV
acr

FACR <- geomean(acr$ACR)
FACR
```

Only `r nrow(acr)` species have both an acute and chronic value in this
filtered dataset, so FACR is a single-species ratio here — a thin basis by
EPA standards, which call for a minimum of three species.

## Step 8 — Final result: CCC

```{r ccc}
CCC <- FAV / FACR

data.frame(quantity = c("FAV (ug/L)", "FACR (unitless)", "CCC = FAV/FACR (ug/L)"),
           value = c(FAV, FACR, CCC))
```

```{r ccc-inline}
#| echo: false
cat(sprintf("CCC = %.4f ug/L microcystin-LR (illustrative only -- see headline conclusion above)\n", CCC))
```

## Limitations

- **Neither dataset meets EPA's 8-family minimum data requirement** — the
  chronic subset has `r nrow(gmcv)` genera / categories, the acute subset
  has `r N_acute` genera / `r length(unique(acute_usable$group))`
  categories. This means the FAV/FACR fallback does not actually resolve
  the data-sufficiency problem; it only relocates it.
- **FACR is based on a single species** (*Daphnia magna*). Restricting to
  apical endpoints flipped its ACR from an atypical value below 1 (in an
  earlier, non-endpoint-filtered version of this analysis) to a more
  typical value above 1 here — a useful illustration of how much a
  single-species ratio can swing with small scoping changes.
- **Acute Values are NOEC/LOEC-based surrogates**, not true LC50/EC50s,
  since the latter weren't consistently reported in this dataset.
- **The FAV formula was not independently re-verified against a freshly
  retrieved primary source** in the session that produced this script (see
  Step 6).
- Excluding all censored (single-bound) rows is conservative for
  transparency but discards real information. A documented variant in
  Step 4 tests including them as surrogate Chronic Values: it just barely
  reaches the 4-GMCV minimum needed to run the FCV formula (FCV ≈ 0.573
  µg/L), but adds no new taxonomic categories, so it doesn't move the
  dataset any closer to being guideline-compliant — see that section for
  the full caveats on why this isn't standard practice.
- **The result is highly scope-sensitive**: this version (acute restricted
  to Mortality only) gives CCC ≈ 0.0277 µg/L; a prior version (acute using
  all four apical endpoints) gave CCC ≈ 0.0235 µg/L; and the earliest
  version with no endpoint filter at all gave CCC ≈ 1.34 µg/L — nearly a
  60-fold range across three reasonable-sounding scoping choices. Treat
  all of these as bracketing examples of method mechanics on thin data,
  not as converging estimates of a true value.

## Reproducing this with different assumptions

To explore sensitivity to the scoping choices made above, edit:

- `KEEP_ENDPOINTS` in **Setup** (e.g., add back Histology or Subcellular
  endpoints).
- The `ca_relevant` filter in **Step 3** (e.g., drop the filter entirely,
  or switch to `ca_native`/`ca_resident`).
- The censoring rule in **Steps 4 and 5** (e.g., keep NOEC-only rows and
  use NOEC itself as `ChV`/`AV` instead of dropping them).
- The `endpoint == "Mortality"` restriction in **Step 5** (e.g., broaden
  back to all four apical endpoints for the acute side, matching the
  chronic side).
