---
title: "Species Sensitivity Distribution (SSD) Construction for Microcystins in California Freshwater Aquatic Species"
subtitle: "EPA (1985) Guidelines methodology"
author: "SCCWRP, Leah Thornton Hampton"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    df-print: kable
  docx:
    toc: true
execute:
  echo: true
  warning: false
---

## Overview

This document constructs Species Sensitivity Distributions (SSDs) for
microcystin toxicity data (`ToxSummary.csv`) using the `ssdtools` R package.
An SSD describes the variation in sensitivity across species by fitting a
statistical distribution — here a log-normal — to single per-species toxicity
values, then reading the 5th percentile (HC5) as the concentration protective
of 95% of species.

Four data-preparation decisions are applied before fitting, in this order:

1. **Endpoint filter:** only apical endpoints (*Growth, Development,
   Reproduction, Mortality,* or *Growth + Development*) are retained.
   Sub-organismal endpoints (Behavior, Histology, Subcellular) are excluded
   because they do not represent whole-organism effects.
2. **NOEC derivation:** where a NOEC is directly reported it is used as-is.
   Where only a LOEC is available, `NOEC = LOEC / 2` is used as a conservative
   surrogate — this follows a common practice of treating the midpoint between
   the assumed true NOEC and the LOEC as the best single estimate.
3. **Acute-to-chronic adjustment:** acute test values are divided by 10 before
   inclusion. This is a simple fixed acute-to-chronic ratio (ACR = 10) used to
   place acute and chronic data on a comparable chronic-equivalent scale,
   consistent with the default ACR commonly applied when species-specific ACRs
   are unavailable.
4. **Per-species collapse:** the median adjusted NOEC across all tests for a
   species is used as that species' single SSD input value, preventing heavily
   studied species from disproportionately influencing the fit.

Four species-scope variants are presented:

| Section | Filter | Rationale |
|---|---|---|
| All Species | none | broadest possible dataset |
| California Relevant | `ca_relevant == "Yes"` | species judged relevant to California conditions (residents + common non-native test species) |
| California Residents | `ca_resident == "Yes"` | species with established populations in California |
| California Native Genus | `ca_native_genus == "Yes"` | species whose genus is native to California, even if the tested species itself is not |

A fifth variant (`ca_native == "Yes"`, strictly native species only) is
attempted but fails due to insufficient data — see the final section.

## Setup

```{r}
#| label: packages
library(tidyverse)
library(ssdtools)
library(ggplot2)

zissou1_extended <- c(
  navy   = "#1c1c84",
  teal   = "#3B9AB2",
  blue   = "#78B7C5",
  yellow = "#EBCC2A",
  gold   = "#E1AF00",
  red    = "#F21A00",
  maroon = "#e86e0b")

```

The `zissou1_extended` palette provides seven visually distinct colors for
differentiating taxonomic groups across plot panels.

## Data Import and Preparation

```{r}
#import
tox_dat <- read.csv(file = "Data/ToxSummary.csv")


#tidy
tox_dat_tidy <- tox_dat %>%
  #convert loec to noec by dividing by 2 if there is no noec, otherwise use noec
  mutate(noec_ssd = case_when(
    !is.na(noec_ug_l) ~ noec_ug_l,
    is.na(noec_ug_l) & !is.na(loec_ug_l) ~ loec_ug_l/2
  )) %>%
  #if acute exposure duration, divide the noec by 10, otherwise leave it along
  mutate(noec_ssd = if_else(chronic_acute == "acute", noec_ssd/10, noec_ssd)) %>%
  #include only higher level endpoints
  filter(endpoint %in% c("Growth", "Development", "Reproduction", "Mortality", "Growth + Development")) %>%
  #mark invasive species
  mutate(Species = if_else(note == "Invasive", paste0(Species, "*"), Species))

```

Invasive species are flagged with an asterisk in the `Species` label so they
are visually distinguishable on the SSD plots without being excluded from the
fit. Whether to exclude invasives from a regulatory SSD is a policy decision
not resolved here.

## SSD Plots

Each plot section below follows the same workflow: filter to the target species
scope → collapse to one median NOEC per species → fit a log-normal distribution
via `ssd_fit_dists()` → generate predicted percentiles with confidence
intervals via `predict()` → plot.

### All Species

The "kitchen sink" variant uses the full endpoint-filtered dataset with no
geographic scope restriction. It represents the most data-rich SSD and serves
as a baseline for comparing how restricting scope shifts the distribution.

```{r}
#kitchen sink
fit_dat <- tox_dat_tidy %>%
    filter(!is.na(noec_ssd)) %>%
    group_by(Species) %>%
    mutate(Conc = median(noec_ssd)) %>% #collapse NOEC by species
    ungroup() %>%
    group_by(Species, Conc, group) %>%
    summarise()

ssd_fits <- ssd_fit_dists(fit_dat, dists = "lnorm")

ssd_pred <- predict(ssd_fits, ci = TRUE)

ssd_plot(fit_dat, ssd_pred,
  label = "Species",
  color = "group",
  xlab = "Concentration (µg/L)",
  ribbon = TRUE,
  label_size = 4,
  hc = NULL
) +
  expand_limits(x = 3000) +
  scale_color_viridis_d() +
  labs(title = "All Species", color = "Taxonomic Group") +
  theme(text = element_text(size = 16))
```

### California Relevant Species

This variant applies the `ca_relevant == "Yes"` filter, which retains species
judged relevant to California aquatic environments — including both residents
and widely used non-native test species (e.g., standard lab strains).
This is the scope most comparable to the CCC derivation in `CCC.qmd`, which
uses the same `ca_relevant` flag.

```{r}
#cali relevant
fit_dat <- tox_dat_tidy %>%
    filter(
      ca_relevant == "Yes",
      !is.na(noec_ssd)) %>%
    group_by(Species) %>%
    mutate(Conc = median(noec_ssd)) %>% #collapse NOEC by species
    ungroup() %>%
    group_by(Species, Conc, group) %>%
    summarise()

ssd_fits <- ssd_fit_dists(fit_dat, dists = "lnorm")

ssd_pred <- predict(ssd_fits, ci = TRUE)

ssd_plot(fit_dat, ssd_pred,
  label = "Species",
  color = "group",
  xlab = "Concentration (µg/L)",
  ribbon = TRUE,
  label_size = 4,
  hc = NULL
) +
  expand_limits(x = 3000) +
  scale_color_manual(values = unname(zissou1_extended)) +
  labs(color = "Taxonomic Group") +
  theme(text = element_text(size = 16), legend.position = "top")
```

### California Residents

The `ca_resident == "Yes"` filter is more restrictive than `ca_relevant`,
retaining only species with established populations in California. Removing
non-native lab strains typically reduces the species count and may shift the
distribution if those strains were among the more or less sensitive entries.

```{r}
#cali residents
fit_dat <- tox_dat_tidy %>%
    filter(
      ca_resident == "Yes",
      !is.na(noec_ssd)) %>%
    group_by(Species) %>%
    mutate(Conc = median(noec_ssd)) %>% #collapse NOEC by species
    ungroup() %>%
    group_by(Species, Conc, group) %>%
    summarise()

ssd_fits <- ssd_fit_dists(fit_dat, dists = "lnorm")

ssd_pred <- predict(ssd_fits, ci = TRUE)

ssd_plot(fit_dat, ssd_pred,
  label = "Species",
  color = "group",
  xlab = "Concentration (µg/L)",
  ribbon = TRUE,
  label_size = 4,
  hc = NULL
) +
  expand_limits(x = 3000) +
  labs(title = "California Residents", color = "Taxonomic Group") +
  theme(text = element_text(size = 16))
```

### California Native Genus

The `ca_native_genus == "Yes"` filter includes species whose **genus** is
native to California, even if the specific test strain or congener tested is
not strictly native. This is a pragmatic expansion of the native-only scope
that recovers test data from closely related taxa when direct native-species
data are scarce.

```{r}
#cali native genus
fit_dat <- tox_dat_tidy %>%
    filter(
      ca_native_genus == "Yes",
      !is.na(noec_ssd)) %>%
    group_by(Species) %>%
    mutate(Conc = median(noec_ssd)) %>% #collapse NOEC by species
    ungroup() %>%
    group_by(Species, Conc, group) %>%
    summarise()

ssd_fits <- ssd_fit_dists(fit_dat, dists = "lnorm")

ssd_pred <- predict(ssd_fits, ci = TRUE)

ssd_plot(fit_dat, ssd_pred,
  label = "Species",
  color = "group",
  xlab = "Concentration (µg/L)",
  ribbon = TRUE,
  label_size = 4,
  hc = NULL
) +
  expand_limits(x = 3000) +
  scale_color_viridis_d() +
  labs(title = "California Native Genus", color = "Taxonomic Group") +
  theme(text = element_text(size = 16))
```

### California Native — Insufficient Data

The strictest scope (`ca_native == "Yes"`, only species themselves native to
California) results in too few data points for `ssd_fit_dists()` to fit a
reliable log-normal distribution. `ssdtools` generally requires a minimum of
six to eight species for a meaningful SSD; the native-only filter falls below
this threshold. The code is retained here for documentation, but its output
should not be interpreted as a valid SSD.

```{r}
#cali_native - FAILS
fit_dat <- tox_dat_tidy %>%
    filter(
      ca_native == "Yes",
      !is.na(noec_ssd)) %>%
    group_by(Species) %>%
    mutate(Conc = median(noec_ssd)) %>% #collapse NOEC by species
    ungroup() %>%
    group_by(Species, Conc, group) %>%
    summarise()

ssd_fits <- ssd_fit_dists(fit_dat, dists = "lnorm")

ssd_pred <- predict(ssd_fits, ci = TRUE)

ssd_plot(fit_dat, ssd_pred,
  label = "Species",
  color = "group",
  xlab = "Concentration (µg/L)",
  ribbon = TRUE,
  label_size = 4,
  hc = NULL
) +
  expand_limits(x = 3000) +
  scale_color_viridis_d() +
  labs(title = "California Native", color = "Taxonomic Group") +
  theme(text = element_text(size = 16))
```

## Limitations

- **LOEC/2 substitution introduces directional uncertainty.** Using half the
  LOEC as a NOEC surrogate assumes the true NOEC lies halfway between zero
  effect and the first observed effect concentration — a reasonable default
  but one that understates sensitivity for species where the true NOEC is
  close to zero, and overstates it where the dose-response is very steep.
- **Fixed ACR of 10 may not reflect species-specific ratios.** The
  acute-to-chronic conversion factor varies widely across taxa and compounds.
  Using a fixed value of 10 introduces uncertainty into the chronic-equivalent
  concentrations derived from acute tests.
- **Log-normal is assumed, not tested.** A single distribution family
  (`lnorm`) is fitted here. Comparing fit across multiple candidate
  distributions (log-logistic, gamma, Weibull) and selecting based on AIC or
  goodness-of-fit is standard SSD practice but is not performed in this
  document.
- **Per-species medians suppress within-species variability.** Collapsing
  multiple tests per species to a single median discards information about
  life-stage and laboratory variation. A species represented by many tests
  under very different conditions may have a median that does not accurately
  represent any one condition.
- **HC5 confidence intervals widen substantially at small N.** The California
  native genus and residents variants include relatively few species, so the
  confidence ribbon around the fitted curve is correspondingly wide. The HC5
  point estimate should be interpreted with caution in those variants.
- **Invasive species are retained in the fit.** Asterisked species are visually
  flagged but not excluded. A regulatory SSD intended to protect native
  communities would typically exclude invasives from the fit, not merely label
  them.

End of Analysis.
